---
title: "SPIGO"
output: html_document
date: "2026-07-22"
---

```{r Packages, include=FALSE}
library(dplyr)
library(ape)
library(tidyverse)
library(forcats)
library(ggtree)
library(babette)
library(ggplot2)
library(table1)
library(phangorn)
library(treeio)
library(ggnewscale)
library(khroma)


setwd("/Users/tddorji/Library/CloudStorage/OneDrive-TheUniversityofMelbourne/Centre_Pathogen_genomics/Vic_IMP/Australian/")
```



```{r Data loading}
global <- read.delim("/Users/tddorji/Library/CloudStorage/OneDrive-TheUniversityofMelbourne/Centre_Pathogen_genomics/Vic_IMP/Australian/Global_IMP_11_09_2026.tsv") |>
  subset(country=="Australia") # |> select(strain, biosample, date, country, region, isolation.source, isolation.type)


amr_all <- readxl::read_xlsx("/Users/tddorji/Library/CloudStorage/OneDrive-TheUniversityofMelbourne/Centre_Pathogen_genomics/Vic_IMP/Australian/combined_reportable_amr_mechanisms.xlsx") |> 
  select(isolate=Isolate, amr="Reportable AMR genes") #|> subset(amr !="CPase_16S_mcr_NEG")

metadata <- readxl::read_xlsx(
  "/Users/tddorji/Library/CloudStorage/OneDrive-TheUniversityofMelbourne/Centre_Pathogen_genomics/Vic_IMP/Victorian/zDates_global/metadata_out/metadata_combined.xlsx") |>
  dplyr::mutate(date = ifelse(is.na(date), date2, date),
    amr_isolate = dplyr::case_when(
      run %in% amr_all$isolate ~ as.character(run),
      biosample %in% amr_all$isolate ~ as.character(biosample),
      TRUE ~ NA_character_)) |>
  dplyr::select(biosample, amr_isolate, date, source,
    instrument = instrument_model_desc) #|>
  #dplyr::left_join(global, by = "biosample") |> 
     #dplyr::filter(!startsWith(biosample, c("SAMN3106599", "SAMN3106604")))


sra <- read.csv("/Users/tddorji/Library/CloudStorage/OneDrive-TheUniversityofMelbourne/Centre_Pathogen_genomics/Vic_IMP/Australian/SRA_AUS.tab", header = TRUE, sep = "\t")

vic_data <- read_csv("/Users/tddorji/Library/CloudStorage/OneDrive-TheUniversityofMelbourne/Centre_Pathogen_genomics/Vic_IMP/Victorian/Vic_IMP.csv") |> select(-c(V4:V10, db_version)) |> select(- "...1")



species <- readxl::read_xlsx("/Users/tddorji/Library/CloudStorage/OneDrive-TheUniversityofMelbourne/Centre_Pathogen_genomics/Vic_IMP/Australian/combined_speciation.xlsx") |>
  dplyr::rename(species = `Match 1 (reads)`) |>
  dplyr::filter(!Isolate %in% c("SRR13501354", "SRR23384557", "SRR23384559", "SRR23384540", "SRR23384721", "SRR9696925",
      "SRR13501351", "SRR23384859", "SRR23384669")) |>
  dplyr::mutate(genus = sub("\\s+.*$", "", trimws(species))) |>
  dplyr::relocate(genus, .after = species) |> select(isolate = Isolate, species, genus) |> 
  left_join(amr_all, by = "isolate") |> left_join(sra, by=c("isolate"="Run")) |> 
  mutate(BioSample = ifelse(is.na(BioSample), isolate, BioSample)) |> left_join(metadata, by=c("isolate"="biosample")) |> 
  left_join(zont_sample_id, by=c("isolate"="V1")) |> 
  dplyr::mutate(platform = dplyr::coalesce(ont, instrument),
                platform = dplyr::recode(platform, "OXFORD_NANOPORE" = "ONT")) |> 
  left_join(mlst, by=c("isolate"="sample")) |> 
   select(-c(instrument, ont, amr1, isolation.source)) |> 
  select(isolate, BioSample, species, genus, scheme, ST, amr, 
    date=date.x, 
    country, region,
    source, type=isolation.type, platform) |> 
  mutate(region=ifelse(isolate %in% vic_data$accession, "Vic", region)) |> 
  mutate(country=ifelse(isolate %in% vic_data$accession, "Australia", country)) |> 
  left_join(select(vic_data, c(accession, travel=Surv_Travel_countries, present=Surv_presentation)), by=c("isolate"="accession")) 
  
  





Aus_run <- read.csv("/Users/tddorji/Library/CloudStorage/OneDrive-TheUniversityofMelbourne/Centre_Pathogen_genomics/Vic_IMP/Australian/report/summary.tsv", header = TRUE, sep = "\t") |> select(-c(QA, is_control, filesize)) |> subset(Depth >=30) |>  
  left_join(sra, by = c("Isolate" = "Run")) |>
  mutate(match_accession = ifelse(
      Isolate %in% metadata$amr_isolate, Isolate, BioSample)) |>
  relocate(c(BioSample, match_accession), .after = Isolate) |> filter(!grepl(":", Species)) |> 
  rename_with(tolower) |> left_join(amr_all, by = "isolate") |> 
  left_join(metadata, by = c("isolate" = "amr_isolate")) 

mlst <- readxl::read_xlsx("/Users/tddorji/Library/CloudStorage/OneDrive-TheUniversityofMelbourne/Centre_Pathogen_genomics/Vic_IMP/Australian/zmlst_combined.xlsx") |> select(sample, scheme=species, ST)

mob_typer <- read.csv(
  "/Users/tddorji/Library/CloudStorage/OneDrive-TheUniversityofMelbourne/Centre_Pathogen_genomics/Vic_IMP/Australian/zCombined_mobtyper.tab", sep = "\t",
  header = TRUE) |>
  dplyr::filter(rep_type.s. != "-") |>
  dplyr::mutate(sample = sub(":.*", "", sample_id)) |>
  dplyr::relocate(sample, .before = sample_id)

contig_report <- read.csv("/Users/tddorji/Library/CloudStorage/OneDrive-TheUniversityofMelbourne/Centre_Pathogen_genomics/Vic_IMP/Australian/zCombined_contigreport.tab", sep = "\t", header = TRUE) |> 
  filter(molecule_type !="chromosome", rep_type.s. != "-") |> mutate(contig = word(contig_id, 1)) |> 
  relocate(contig, .after = contig_id) |> rename_with(tolower)

imp_plasmid <- read.csv("/Users/tddorji/Library/CloudStorage/OneDrive-TheUniversityofMelbourne/Centre_Pathogen_genomics/Vic_IMP/Australian/imp_plasmid_summary.tsv", sep = "\t", header = TRUE) |> select(1:2) |> 
   mutate(imp_contig = str_trim(imp_contig)) %>%
  left_join(contig_report, by = c("sample" = "sample_id")) |> 
  subset(imp_contig %in% contig) 

```







### Figures

```{r IMP-fig-species}


species_colors <- c(
  "Klebsiella pneumoniae"   = "#0072B2",
  "Escherichia coli"        = "#E69F00",
  "Enterobacter hormaechei" = "#009E73",
  "Serratia marcescens"     = "#CC79A7",
  "Pseudomonas aeruginosa"  = "#D55E00",
  "Citrobacter tructae"     = "#00BFC4",
  "Klebsiella variicola"    = "#F0E442",
  "Klebsiella michiganensis" = "#CC2F8A",
  "Klebsiella oxytoca"      = "#4B0082",
  "Enterobacter asburiae"   = "#8C510A",
  "Citrobacter freundii"    = "#7F7F7F",
  "Klebsiella aerogenes"    = "#56B4E9",
  "Proteus mirabilis"       = "#003366")


species_counts <- species |>
  count(species, sort = TRUE) |>
  slice_max(n, n = 10, with_ties = FALSE) |>
  mutate(prop = n / nrow(Aus_run),
    species = forcats::fct_reorder(species, n))

fig_species <- ggplot(species_counts, aes(x = n, y = species, fill = species)) +
  geom_col() +
  geom_text(aes(label = scales::percent(prop, accuracy = 0.1)),
    hjust = -0.1, size = 3.2) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.15))) +
  scale_fill_manual(values = species_colors, guide = "none") +
  labs(x = "Number of isolates", y = NULL,
    title = "Top 10 IMP-CPO species") +
  theme_classic() +
  theme(axis.text = element_text(size = 13),
    plot.title = element_text(size = 14, hjust = 0.5),
    plot.title.position = "plot")

fig_species

```


```{r Mashtree}
mash <- read.tree("/Users/tddorji/Library/CloudStorage/OneDrive-TheUniversityofMelbourne/Centre_Pathogen_genomics/Vic_IMP/All/mash_imp.dnd")

p <- ggtree(midpoint(mash)) 

tree_data <- data_clean |> select(accession, species, imp_variant, co_genes,year) |> column_to_rownames(var = "accession")

tree_species <- tree_data |> select(species)
tree_variant <- tree_data |> select(imp_variant)
tree_genes   <- tree_data |> select(co_genes)
tree_year    <- tree_data |> select(year) |> mutate(year=as.factor(year))


##
colnames(tree_species) <- "Species"
colnames(tree_variant) <- "IMP variant"
colnames(tree_genes) <- "Co-genes"
colnames(tree_year) <- "Year"

p1 <- gheatmap(p, tree_species,
               offset = 0.001, width = 0.03,
               colnames_angle = 90, colnames_offset_y = -19,
               hjust = 0, font.size = 5) +
    scale_fill_manual(values = species_colors, name = "Species") +
    new_scale_fill()


p2 <- gheatmap(p1, tree_variant,
         offset = 0.009, width = 0.03,
         colnames_angle = 90, 
         colnames_offset_y = -19, 
         hjust = 0, 
         font.size = 5) + 
         ggsci::scale_fill_lancet(name="IMP variants")+
        new_scale_fill()

p3 <- gheatmap(p2, tree_genes,
               offset = 0.017, width = 0.03,
               colnames_angle = 90, 
               colnames_offset_y = -19,
               hjust = 0, font.size = 5) +
    ggsci::scale_fill_bmj(name="Co-genes") + new_scale_fill()

p3

p4 <- gheatmap(p3, tree_year,
               offset = 0.025, 
               width = 0.03,
               colnames_angle = 90, 
               colnames_offset_y = -19,
               hjust = 0, font.size = 5) +
    ggsci:::scale_fill_igv( name = "Year")

p_heatmap <- p4 +
    theme(
        legend.position = "right",
        legend.title = element_text(size = 10, face = "bold"),
        legend.text = element_text(size = 8),
        plot.margin = margin(t = 10, r = 60, b = 40, l = 10)) +
    ylim(NA, max(p$data$y) * 1.15)
p_heatmap

```

```{r IMP-fig-temporal}

data_clean <- Aus_run |>
    select(year, amr) |>
    mutate(imp_variant = stringr::str_extract(amr, "IMP-[0-9]+"))

imp_colors <- c(
  "IMP-4"  = "#1B9E77",
  "IMP-26" = "#D95F02",
  "IMP-7"  = "#7570B3",
  "IMP-1"  = "#E7298A",
  "IMP-62" = "#66A61E",
  "IMP-14" = "#E6AB02",
  "IMP-15" = "#A6761D")

fig_temporal <- ggplot(data_clean, aes(x = year, fill = imp_variant)) +
  geom_bar() +
  scale_x_continuous(breaks = scales::breaks_width(1)) +
  scale_fill_manual(values = imp_colors) +
  labs(x = "Year of collection",
    y = "Number of isolates",
    fill = "IMP variant") + 
  theme_classic() +
  theme(plot.title.position = "plot",
    plot.title = element_text(hjust = 0.5))

fig_temporal

```

```{r IMP species temporal}

species <- Aus_run |>
  count(year, species, name = "n")

fig_variant <- ggplot(species, aes(x = factor(year), y = n, fill = species)) +
  geom_col() +
  scale_fill_manual(values = species_colors) +
  labs(x = "Year", y = "Number of isolates", fill = "Species") +
  theme_classic()

fig_variant
```

```{r IMP-fig-st}

fig_st <- ST |>
  slice_max(n_typed, n = 10, with_ties = FALSE) |>
  ggplot(aes(x = n_typed, y = n_STs, color = species)) +
  geom_abline(
    slope = 1, intercept = 0,
    linetype = "dashed", color = "grey60") +
  geom_point(size = 3) +
  scale_color_manual(values = species_colors) +
  labs(
    x = "Isolates with ST assigned",
    y = "Distinct STs",
    color = "Species",
    title = "Sequence type diversity: top 10 species",
    subtitle = "Points below the dashed line indicate repeated STs") +
  theme_classic()

fig_st
```

```{r IMP-fig-presentation}

fig_presentation <- ggplot(data_clean, aes(y = fct_rev(species), fill = Presentation)) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = presentation_col) +
  labs(x = "Proportion of isolates", y = NULL, fill = "Clinical presentation",
       title = "Clinical presentation by species") + theme_classic()
fig_presentation

```

### Statistical analysis

Species has many levels with small counts, so species with fewer than 5
isolates are lumped into "Other (n<5 each)" before testing, to reduce (but not
eliminate) sparse cells. Expected cell counts are well below 5 in most cells of
the raw 13x3 table, so a chi-square test is not valid here; Fisher's exact
test (Monte Carlo p-value, since the table is too large for exact enumeration)
is used instead for both comparisons below. Given only two planned tests, no
multiple-testing correction is applied, but results should be treated as
hypothesis-generating given the sparse cell counts.

```{r IMP-stats-presentation}
data_test <- data_clean |>
  mutate(Species_grp = fct_lump_min(species, min = 5, other_level = "Other (n<5 each)"))

tab_presentation <- table(data_test$Species_grp, data_test$Presentation)
tab_presentation

fisher_presentation <- fisher.test(tab_presentation, simulate.p.value = TRUE, B = 10000)
fisher_presentation

chi_presentation <- suppressWarnings(chisq.test(tab_presentation, simulate.p.value = TRUE, B = 10000))
cramers_v_presentation <- sqrt(chi_presentation$statistic / (sum(tab_presentation) * (min(dim(tab_presentation)) - 1)))
cramers_v_presentation
```



```{r IMP-stats-variant}
data_test <- data_test |>
  mutate(imp4 = if_else(imp_variant == "IMP-4", "IMP-4", "Other IMP variant"))

tab_variant <- table(data_test$Species_grp, data_test$imp4)
tab_variant

fisher_variant <- fisher.test(tab_variant, simulate.p.value = TRUE, B = 10000)
fisher_variant
```

IMP variant is strongly associated with species (Fisher's exact Monte Carlo
p = `r signif(fisher_variant$p.value, 2)`): every species except
*Pseudomonas aeruginosa* carries almost exclusively IMP-4 (117/122 non-*P.
aeruginosa* isolates), whereas *P. aeruginosa* carries a mix of variants
(IMP-7, IMP-1, IMP-14, IMP-15, IMP-26, IMP-62; only 3/11 IMP-4). This likely
reflects species-associated mobile genetic element/integron backgrounds rather
than a direct biological interaction between species and enzyme variant, and
should be interpreted descriptively.



```{r AMR genes location}

amr_loc = read.delim("/Users/tddorji/Library/CloudStorage/OneDrive-TheUniversityofMelbourne/Centre_Pathogen_genomics/Vic_IMP/All/zall_mobsuite_IMP_contig_report.txt")

```


```{r Enterobacter Hormacgei}

entero <- species |> subset(species=="Enterobacter hormaechei") 




```
